On $k$-tactics in Gruenhage's compact-point game

### 2015-03-27 articles/k-tactics-in-gruenhages-compact-point-game/article.tex
incorporate suggestions from G

\begin{abstract}
Gary Gruenhage showed in \cite{MR858337} that metacompactness and
$\sigma$-metacompactness may be characterized for locally compact spaces
by way of a certain topological game using limited information strategies
which consider only the most recent move of the opponent (tactical/stationary
strategies) and possibly the round number (Markov strategies). This paper
demonstrates a non-trivial example of a space for which a winning strategy
exists but every limited information strategy considering a maximum of $k$
moves of the opponent and the round number may be defeated. The question
follows: are metacompactness and $\sigma$-metacompactness characterized by
winning $k$-tactical and $k$-Markov strategies?
\end{abstract}

@@

\begin{game}
  Let $\gruKPGame{X}$ denote the \term{Gruenhage compact/point game}
  with players $\pl K$, $\pl P$. During round $n$, $\pl K$ chooses
  a compact subset $K_n$ of $X$, followed by $\pl P$ choosing a point
  $p_n\in X$ such that $p_n\not\in \bigcup_{m\leq n}K_m$.

@@

\begin{definition}
  A \term{strategy} for a game $G$ with moveset $M$ is a function
  $\sigma:M^{<\omega}\to M$; intuitively, this is a fixed rule for one player's
  choices in consideration of her opponent's moves. If using such a strategy
  always results in a win for the player $\pl P$ using it, then it is called a
  \term{winning strategy} and we write $\pl P\win G$.
\end{definition}

@@

The above game was used by the author's adviser Gary Gruenhage in
\cite{MR858337} to characterize
metacompactness (every open cover of the space has a point-finite open
refinement covering the space) and $\sigma$-metacompactness
(every open cover of the space has a $\sigma$-point-finite open refinement
covering the space) in locally compact spaces. However, the characterization
considers so-called \term{limited information strategies} which do not use
full information of the history of the game.

\begin{definition}
  A $k$-tactical strategy for a game $G$ with moveset $M$ is a function
  $\sigma:M^{\leq k}\to M$; intuitively, it is a strategy which only considers
  the previous $k$ moves of the opponent. If using such a strategy
  always results in a win for the player $\pl P$ using it, then
  we write $\pl P\ktactwin{k} G$.
\end{definition}

\begin{definition}
  A $k$-Markov strategy for a game $G$ with moveset $M$ is a function
  $\sigma:M^{\leq k}\times\omega\to M$; intuitively, it is a strategy which
  only considers the previous $k$ moves of the opponent and the round number.
  If using such a strategy
  always results in a win for the player $\pl P$ using it, then
  we write $\pl P\kmarkwin{k} G$.
\end{definition}

@@

For instance, suppose $\pl P$ wants to use a winning $2$-tactic $\sigma$ from
$\gruKPGame{X}$ to create a winning $2$-tactic for
$\gruConGame{\onePtComp X}{\infty}$. If $\pl P$ is attacked by
$\<x_0,\infty,x_2,\infty,\dots\>$, then $\pl P$ must ensure that the $x_{2n}$
converge without using the point $\infty\not\in X$.
It's difficult to see how, as $\pl P$ may not take advantage of
$\sigma(\<x_{2n},x_{2n+2}\>)$;
$x_{2n},x_{2n+2}$ are not consecutive moves in the original game.

@@

\begin{definition}
  Let $\pmb X=(X\times 2^{<\omega})\cup C$ denote a Cantor tree
  of copies of a zero-dimensional, compact space $X$ with a point-countable
  cover $\mc{U} = \{U_\alpha : \alpha<\omega_1\}$ of distinct
  clopen sets, along with an uncountable subset of the Cantor set
  $C=\{c_\alpha:\alpha<\omega_1\}\in [2^\omega]^{\omega_1}$.
  The topology on $\pmb X$ is given by decaring $U\times\{s\}$ to be a open
  neighborhood of $\<x,s\>\in X\times 2^{<\omega}$ for each
  open neighborhood $U$ of $x$ in $X$, and declaring
  $
    B_{\alpha,m}
      =
    (U_\alpha\times\{c_\alpha\rest n: m\leq n<\omega\})\cup\{c_\alpha\}
  $
  to be a clopen neighborhood of $c_\alpha\in C$ for each $\alpha<\omega_1$,
  $m<\omega$.
\end{definition}

@@

Figure \ref{fig:pmbX} provides a rough illustration of
$\pmb X$, $L_m$ (the triangle at the base), and $K_F'$
(the branches of open sets descending from $C$).

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{pmbX.pdf}
  \caption{$\pmb X$, with $L_m$ and $K_F'$}
  \label{fig:pmbX}
\end{figure}

\begin{lemma}
  $K_F$, $K_F'$, and $L_m$ are compact in $\pmb X$.
  Furthermore, every compact set is contained in a union of $K_F'$, $L_m$
  for some $F\in C^{<\omega}$ and $m<\omega$.
\end{lemma}

@@

\begin{theorem}
  $\pl K\win\gruKPGame{\pmb X}$.
\end{theorem}

@@

  Let $M:\pmb X\times \omega\to \mc K(\pmb X)$ as follows:
    \[
      M(\pmb x,n)
        =
      \left\{
        \begin{array}{lcl}
          K_{\{\alpha_{x,m}:m\leq n\}}
        & : &
          \pmb x = \<x,s\> \in X\times 2^{<\omega}
        \\
          K_{\{\alpha\}}
        & : &
          \pmb x = c_\alpha \in C
        \end{array}
      \right.
    \]
  and use $M$ to define the strategy $\sigma$ for each
  $\pmb a\in\pmb X^{<\omega}$:
    \[
      \sigma(\pmb a)
        =
      L_{|\pmb a|}
        \cup
      \bigcup_{i< |\pmb a|}
      M(\pmb a(i),|\pmb a|)
    \]

  Let $\pmb p:\omega\to\pmb X$ be a legal attack against $\sigma$. Then as
  $\pmb p(n)\not\in L_n$, for each $\pmb x=\<x,s\>\in X\times 2^{<\omega}$,
  $X\times\{s\}$ is an open neighborhood of $\pmb x$ which contains finitely
  many $\pmb p(n)$.

@@

One might hope then that $\pl K\nottactwin\gruKPGame{\pmb X}$ but
$\pl K\ktactwin{2}\gruKPGame{\pmb X}$, giving us our counterexample.
However, we will see that in fact, any winning $k$-mark for $\pl K$ may
be improved to a winning tactic by exploiting the structure of the Cantor tree.

In particular, knowledge of round number does not assist $\pl K$,
since she may force $\pl P$ to either stay within $C$, or to seed a growing
integer which could be used in place of the round number
by forcing her to play outside $L_{|s|+1}$ in response to
$\<x,s\>\in X\times 2^{<\omega}$.

\begin{lemma}
  If $\pl K\kmarkwin{(k+1)}\gruKPGame{\pmb X}$, then
  $\pl K\ktactwin{(k+1)}\gruKPGame{\pmb X}$.
\end{lemma}

@@

  Define $r:\pmb X\to\omega$ by
    \[
      r(\pmb x)
        =
      \left\{
        \begin{array}{lcl}
          |s|
        & : &
          \pmb x = \<x,s\> \in X\times 2^{<\omega}
        \\
          0
        & : &
          \pmb x \in C
        \end{array}
      \right.
    \]
  and use $r$ to define the $(k+1)$-tactic $\tau$ by
    \[
      \tau(\emptyset)
        =
      \sigma(\emptyset,0)
    \]
    \[
      \tau(\pmb t\concat\<\pmb x\>)
        =
      L_{r(\pmb x)+1}
        \cup
      \{\pmb x\}
        \cup
      \sigma(\pmb t\concat\<\pmb x\>,r(\pmb x)+1)
    \]

  Let $\pmb p:\omega\to\pmb X$ be a legal attack by $\pl P$ against $\tau$.
  If $\pmb p(n)\in C$ for $N<n<\omega$, then since no $\pmb p(n)$ may be legally
  repeated, $\{\{\pmb p(n)\}:N<n<\omega\}$ is a discrete collection, making
  the points $p(n)$ locally finite.

@@

\begin{corollary}
  $\pmb X$ is $\sigma$-metacompact if and only if $\pmb X$ is metacompact.
\end{corollary}

@@

\begin{lemma}
  If $\pl K\ktactwin{(k+1)}\gruKPGame{\pmb X}$, then
  $\pl K\tactwin\gruKPGame{\pmb X}$.
\end{lemma}

@@

  Let $\pmb p:\omega\to \pmb X$ be a legal attack against
  $\tau$, and assume $\pmb p(n)=\<x_n,s_n\>\in X\times 2^{<\omega}$.
  Then consider the attack $\pmb q:\omega\to X\times 2^{<\omega}$ against
  $\sigma$ defined by, for $n<\omega$ and $i<k$,
    \[
      \pmb q((k+1)n)=\pmb p(n)=\<x_n,s_n\>
    \]
    \[
      \pmb q((k+1)n+(i+1))=\<y, a_{M^{i+1}(x_n,|s_n|)}\>
    \]

@@

So our hopes for a counterexample to our main question
in $\pmb X$ are thusly defeated. Our
consolation prize will be to show that for a certain choice of $X$
and $\{U_\alpha:\alpha<\omega_1\}$, $\pmb X$ is at least
an example of a space which allows a winning strategy for $\gruKPGame{\pmb X}$
but no winning $k$-tactics or $k$-marks.

To this end, we will show that $\{U_\alpha:\alpha<\omega_1\}$ may be chosen
such that $\pmb X$ is not metacompact, and therefore
$\pl K$ lacks a $k$-Markov strategy for any $k<\omega$.

\begin{theorem}
  $\pmb X$ is metacompact
    if and only if
  $\{U_\alpha:\alpha<\omega_1\}$ is $\sigma$-point-finite.
\end{theorem}

@@

  Let $\mc U$ be a cover of $\pmb X$, and for each $s\in 2^\omega$
  let $\mc V_s$ be a finite open refinement of $\mc C$ covering the compact
  set $X\times\{s\}$. Then let $\mc W_n = \{B_{\alpha,n_\alpha}:\alpha\in A_n\}$
  be an open refinement of $\mc C$ for each $n<\omega$, and note that it
  is point-finite. It follows that
  $\mc U' = \bigcup_{s\in 2^{<\omega}} \mc V_s \cup \bigcup_{n<\omega} \mc W_n$
  is an open $\sigma$-point-finite refinement of $\mc U$, so $\pmb X$
  is $\sigma$-metacompact, and therefore it is metacompact.

  For the other direction, consider the open cover
  $\mc U = \{B_{\alpha,0}:\alpha<\omega_1\}$ of the closed subset $C$ of
  metacompact $\pmb X$, and let $\{B_{\alpha,n_\alpha}:\alpha<\omega_1\}$ be
  a point-finite refinement. Then
  $\mc U_s = \{ U_\alpha : c_\alpha \rest n_\alpha = s\}$ is
  point-finite for $s\in 2^{<\omega}$ and
  $U_\alpha\in\mc U_{c_\alpha\rest n_\alpha}$ for each $\alpha<\omega_1$.
  Therefore $\mc U = \bigcup_{s\in 2^{<\omega}} \mc U_s$ is
  $\sigma$-point-finite.
\end{proof}

>>>>

\begin{abstract}
Gary Gruenhage showed in \cite{MR858337} that metacompactness and\break
$\sigma$-metacompactness may be characterized for locally compact spaces
by way of a certain topological game using limited information strategies:
strategies
which consider only the most recent move of the opponent (tactical/stationary
strategies) and possibly the round number (Markov strategies). This paper
demonstrates a non-trivial example of a space for which a winning strategy
exists but every limited information strategy considering a maximum of $k$
moves of the opponent and the round number may be defeated. The question
follows: are metacompactness and $\sigma$-metacompactness characterized by
winning $k$-tactical and $k$-Markov strategies for $k>1$?
\end{abstract}

@@

\begin{game}
  Let $\gruKPGame{X}$ denote \term{Gruenhage's compact/point game}
  with players $\pl K$, $\pl P$. During round $n$, $\pl K$ chooses
  a compact subset $K_n$ of $X$, followed by $\pl P$ choosing a point
  $p_n\in X$ such that $p_n\not\in \bigcup_{m\leq n}K_m$.

@@

\begin{definition}
  A \term{strategy} for a game $G$ with moveset $M$ is a function
  $\sigma:M^{<\omega}\to M$; intuitively, this is a fixed rule for one player's
  choices in consideration of her opponent's moves. If using such a strategy
  always results in a win for the player $\pl P$ using it, then it is called a
  \term{winning strategy}. If a winning strategy exists for $\pl P$ in the
  game $G$, then we write $\pl P\win G$.
\end{definition}

@@

$\gruKPGame{X}$ was used by Gary Gruenhage in
\cite{MR858337} to characterize
metacompactness (every open cover of the space has a point-finite open
refinement covering the space) and $\sigma$-metacompactness
(every open cover of the space has a $\sigma$-point-finite open refinement
covering the space) in locally compact spaces. This characterization
considers so-called \term{limited information strategies} which do not use
full information of the history of the game.

\begin{definition}
  A $k$-tactical strategy for a game $G$ with moveset $M$ is a function
  $\sigma:M^{\leq k}\to M$; intuitively, it is a strategy which only considers
  the previous $k$ moves of the opponent. If a winning $k$-tactical strategy
  exists for $\pl P$ in the game $G$, then we write $\pl P\ktactwin{k} G$.
\end{definition}

\begin{definition}
  A $k$-Markov strategy for a game $G$ with moveset $M$ is a function
  $\sigma:M^{\leq k}\times\omega\to M$; intuitively, it is a strategy which
  only considers the previous $k$ moves of the opponent and the round number.
  If a winning $k$-Markov strategy exists for
  $\pl P$ in the game $G$, then we write $\pl P\kmarkwin{k} G$.
\end{definition}

@@

For instance, suppose $\pl O$ wants to use a winning $2$-tactic $\sigma$
for $\pl K$ in
$\gruKPGame{X}$ to create a winning $2$-tactic for
$\gruConGame{\onePtComp X}{\infty}$. If $\pl O$ is attacked by
$\<x_0,\infty,x_2,\infty,\dots\>$, then $\pl O$ must ensure that the $x_{2n}$
converge without considering the point $\infty\not\in X$.
It's difficult to see how, as $\pl O$ may not take advantage of
$\sigma(\<x_{2n},x_{2n+2}\>)$;
$x_{2n},x_{2n+2}$ are not consecutive moves in the original game.

@@

\begin{definition}
  Let $\mb X=(X\times 2^{<\omega})\cup C$ denote a Cantor tree
  of copies of a zero-dimensional, compact space $X$ with a point-countable
  cover $\mc{U} = \{U_\alpha : \alpha<\omega_1\}$ of distinct
  clopen sets, along with an uncountable subset of the Cantor set
  $C=\{c_\alpha:\alpha<\omega_1\}\in [2^\omega]^{\omega_1}$.
  The topology on $\mb X$ is given by decaring $U\times\{s\}$ to be a open
  neighborhood of $\<x,s\>\in X\times 2^{<\omega}$ for each
  open neighborhood $U$ of $x$ in $X$, and declaring
  $
    B_{\alpha,m}
      =
    (U_\alpha\times\{c_\alpha\rest n: m\leq n<\omega\})\cup\{c_\alpha\}
  $
  to be a clopen neighborhood of $c_\alpha\in C$ for each $\alpha<\omega_1$,
  $m<\omega$.
\end{definition}

@@

Figure \ref{fig:pmbX} provides a rough illustration of
$\mb X$, $L_m$ (the triangle at the base), and $K_F'$
(the branches of open sets descending from $C$).

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{pmbX.pdf}
  \caption{$\mb X$, with $L_m$ and $K_F'$}
  \label{fig:pmbX}
\end{figure}

\begin{lemma}
  $K_F$, $K_F'$, and $L_m$ are compact in $\mb X$.
  Furthermore, every compact set is contained in a union of $K_F'$, $L_m$
  for some $F\in C^{<\omega}$ and $m<\omega$.
\end{lemma}

@@

\begin{theorem}
  $\pl K\win\gruKPGame{\mb X}$.
\end{theorem}

@@

  Let $M:\mb X\times \omega\to \mc K(\mb X)$ as follows:
    \[
      M(\pmb x,n)
        =
      \left\{
        \begin{array}{lcl}
          K_{\{\alpha_{x,m}:m\leq n\}}
        & : &
          \pmb x = \<x,s\> \in X\times 2^{<\omega}
        \\
          K_{\{\alpha\}}
        & : &
          \pmb x = c_\alpha \in C
        \end{array}
      \right.
    \]
  and use $M$ to define the strategy $\sigma$ for each
  $\pmb a\in\mb X^{<\omega}$:
    \[
      \sigma(\pmb a)
        =
      L_{|\pmb a|}
        \cup
      \bigcup_{i< |\pmb a|}
      M(\pmb a(i),|\pmb a|)
    \]

  Let $\pmb p:\omega\to\mb X$ be a legal attack against $\sigma$. Then as
  $\pmb p(n)\not\in L_n$, for each $\pmb x=\<x,s\>\in X\times 2^{<\omega}$,
  $X\times\{s\}$ is an open neighborhood of $\pmb x$ which contains finitely
  many $\pmb p(n)$.

@@

One might hope then that $\pl K\nottactwin\gruKPGame{\mb X}$ but
$\pl K\ktactwin{2}\gruKPGame{\mb X}$, giving us our counterexample.
However, we will see that in fact, any winning $k$-tactic or even
$k$-mark for $\pl K$ may
be improved to a winning tactic by exploiting the structure of the Cantor tree.

We first show that knowledge of round number does not assist $\pl K$,
since she may force $\pl P$ to either stay within $C$, or to seed a growing
integer which could be used in place of the round number
by forcing her to play outside $L_{|s|+1}$ in response to
$\<x,s\>\in X\times 2^{<\omega}$.

\begin{lemma}
  If $\pl K\kmarkwin{(k+1)}\gruKPGame{\mb X}$, then
  $\pl K\ktactwin{(k+1)}\gruKPGame{\mb X}$.
\end{lemma}

@@

  Define $r:\mb X\to\omega$ by
    \[
      r(\pmb x)
        =
      \left\{
        \begin{array}{lcl}
          |s|
        & : &
          \pmb x = \<x,s\> \in X\times 2^{<\omega}
        \\
          0
        & : &
          \pmb x \in C
        \end{array}
      \right.
    \]
  and use $r$ to define the $(k+1)$-tactic $\tau$ by
    \[
      \tau(\emptyset)
        =
      \sigma(\emptyset,0)
    \]
    \[
      \tau(\pmb t\concat\<\pmb x\>)
        =
      L_{r(\pmb x)+1}
        \cup
      \{\pmb x\}
        \cup
      \sigma(\pmb t\concat\<\pmb x\>,r(\pmb x)+1)
    \]

  Let $\pmb p:\omega\to\mb X$ be a legal attack by $\pl P$ against $\tau$.
  If $\pmb p(n)\in C$ for $N<n<\omega$, then since no $\pmb p(n)$ may be legally
  repeated, $\{\{\pmb p(n)\}:N<n<\omega\}$ is a discrete collection, making
  the points $p(n)$ locally finite.

@@

\begin{corollary}
  $\mb X$ is $\sigma$-metacompact if and only if $\mb X$ is metacompact.
\end{corollary}

@@

\begin{lemma}
  If $\pl K\ktactwin{(k+1)}\gruKPGame{\mb X}$, then
  $\pl K\tactwin\gruKPGame{\mb X}$.
\end{lemma}

@@

  Let $\pmb p:\omega\to \mb X$ be a legal attack against
  $\tau$, and assume $\pmb p(n)=\<x_n,s_n\>\in X\times 2^{<\omega}$.
  Then consider the attack $\pmb q:\omega\to X\times 2^{<\omega}$ against
  $\sigma$ defined by, for $n<\omega$ and $i<k$,
    \[
      \pmb q((k+1)n)=\pmb p(n)=\<x_n,s_n\>
    \]
    \[
      \pmb q((k+1)n+(i+1))=\<y, a_{M^{i+1}(x_n,|s_n|)}\>
    \]

@@

So our hopes for a counterexample to our main question
in $\mb X$ are thusly defeated. Our
consolation prize will be to show that for a certain choice of $X$
and $\{U_\alpha:\alpha<\omega_1\}$, $\mb X$ is at least
an example of a space which allows a winning strategy for $\gruKPGame{\mb X}$
but no winning $k$-tactics or $k$-marks.

To this end, we will show that $\{U_\alpha:\alpha<\omega_1\}$ may be chosen
such that $\mb X$ is not metacompact, and therefore
$\pl K$ lacks a $k$-Markov strategy for any $k<\omega$.

\begin{theorem}
  $\mb X$ is metacompact
    if and only if
  $\{U_\alpha:\alpha<\omega_1\}$ is $\sigma$-point-finite.
\end{theorem}

@@

  Let $\mc U$ be a cover of $\mb X$, and for each $s\in 2^\omega$
  let $\mc V_s$ be a finite open refinement of $\mc C$ covering the compact
  set $X\times\{s\}$. Then let $\mc W_n = \{B_{\alpha,n_\alpha}:\alpha\in A_n\}$
  be an open refinement of $\mc C$ for each $n<\omega$, and note that it
  is point-finite. It follows that
  $\mc U' = \bigcup_{s\in 2^{<\omega}} \mc V_s \cup \bigcup_{n<\omega} \mc W_n$
  is an open $\sigma$-point-finite refinement of $\mc U$, so $\mb X$
  is $\sigma$-metacompact, and therefore it is metacompact.

  For the other direction, consider the open cover
  $\mc U = \{B_{\alpha,0}:\alpha<\omega_1\}$ of the closed subset $C$ of
  metacompact $\mb X$, and let $\{B_{\alpha,n_\alpha}:\alpha<\omega_1\}$ be
  a point-finite refinement. Then
  $\mc U_s = \{ U_\alpha : c_\alpha \rest n_\alpha = s\}$ is
  point-finite for $s\in 2^{<\omega}$ and
  $U_\alpha\in\mc U_{c_\alpha\rest n_\alpha}$ for each $\alpha<\omega_1$.
  Therefore $\mc U = \bigcup_{s\in 2^{<\omega}} \mc U_s$ is
  $\sigma$-point-finite.
\end{proof}
